\chapter{Acknowledgements}

Thanks must go first to my primary supervisor, Peter Cox. His willingness to support me and my ideas, combined with his own insight
have been incredibly useful over these years. My second supervisor Peter Ashwin also deserves my thanks,
particularly for his ability to help me understand technical topics.

I should also thank my co-authors, Chris Huntingford and Paul Ritchie. The time you took to help me
write my papers, from the drafting stage right through to answering reviewer comments was always greatly appreciated. Paul
in particular deserves my thanks for his role as (amongst other things) my unofficial supervisor. One day I will beat him at squash.

I was fortunate to work in the supportive environment of Exeter Climate Systems (XCS). I would like to thank in particular
all those in Office 901, old and new, who created such a friendly place to work. I started in The Dungeon,
but eventually I was allowed my own desk (with a whiteboard!).

Those friends I made along the way, whether through a Laver lunch on the balcony or, more probably, by spending time at Climate and
Friends have also earned my thanks. Through trips to the beach, Dartmoor and even further abroad, they have reminded me that there is more to life
than bifurcation diagrams.

Ellie and Margaret, who I live with, also have my gratitude. The pandemic meant we saw more of each other than any of us wanted.
Thank you for all the cups of tea and trips to the `pub' which kept me sane over lockdown. I am sorry for
all the food I stole.

I would also like to thank my family for all their love and support over the years, not least for taking me in over the first lockdown.
It was a strange and disorienting time but being at home made it much more bearable.

Finally, I would like to thank Alice (and Milo), who have made these last few years so happy.